Paper 1: Instrument development

### 2024-09-02 Rmd/Paper1.Rmd
Added checks for outliers and missing

@@

@@

@@

Missing data reduces the size of the dataset used for analysis, which leads to lower power (i.e., larger chances for Type II-error, that is not detecting significant effects, obtaining false negatives). There are different advice for how to handle missing data, depending on whether they are missing at random (MAR), completely at random (MCAR) or not at random (NMAR) [@rubin_inference_1975; @schafer_missing_2002]. Data that are missing at random are correlated with the questionnaire (i.e., a question may be missing from a specific version of a questionnaire, which is in a sense random). Data that are missing completely at random, do however not correlate with anything obvious. They are in a sense, seemingly truly random, with no clear rationale for their missing other than circumstance. Data missing not at random are questions that correlate with the question itself (i.e., several respondents may avoid a specific question, i.e., about gender, there seems to be a rationale, that is not a random cause, for data missing for this specific item).

@@

>>>>

library(psych)                              # for skew
library(mice)                               # for missing

@@

```{r}
### HELP FUNCTION
# describe data (mean, sd, skew, curtosis)
# requires library(moments) for skew
describe_data <- function (datalist, QUIET=1) {

  im  <- mean(datalist, na.rm=TRUE)
  isd <- sd(datalist, na.rm=TRUE)
  isk <- psych::skew(datalist, na.rm=TRUE) # from psych
  ik  <- kurtosi(datalist, na.rm=TRUE) # from psych
  mis <- sum(is.na(datalist)) # missing
  ir  <- max(datalist, na.rm=TRUE) - min(datalist, na.rm=TRUE) # range

  # source: https://stackoverflow.com/questions/34351598/getting-name-of-variable-in-r
  if (QUIET != 1) {
    print(sprintf("%7s = %s",   "Item",    deparse(substitute(datalist))))
    print(sprintf("%7s = %.3f", "mean",    im))
    print(sprintf("%7s = %.3f", "std.dev", isd))
    print(sprintf("%7s = %.3f", "skew",    isk))
    print(sprintf("%7s = %.3f", "kurt",    ik))
    print(sprintf("%7s = %.3f", "range",   ir))
    print(sprintf("%7s = %i",   "missing", mis))

    if (isk > 2 | ik > 7) {
      print("                            Skew or curtosis TOO LARGE.")
    }
  }

  # outliers
  lowerq = quantile(datalist, na.rm=TRUE)[2]
  upperq = quantile(datalist, na.rm=TRUE)[4]
  iqr    = upperq - lowerq               #Or use IQR(df)
  # gang med 3 for extreme outliers, 1.5 for mild outliers

  mild.threshold.upper = (iqr * 3) + upperq
  mild.threshold.lower = lowerq - (iqr * 3)

  # print outliers
  # kilde: https://stackoverflow.com/questions/19353116/how-do-i-print-values-in-a-list-that-are-greater-than-a-certain-number-along-wit#19353192
  outliers_upper <- which(datalist>mild.threshold.upper)
  outliersu      <- datalist[outliers_upper]
  outliers_lower <- which(datalist<mild.threshold.lower)
  outliersl      <- datalist[outliers_lower]

  if (length(outliersu) > 0 | length(outliersl) > 0) {
    print("outliers upper =")
    print(outliersu)
    print("outliers lower =")
    print(outliersl)
  }

  #hist (datalist, main=i) # histogram
  #readline(prompt = "Press enter")

  return(c("mean"=round(im,3), "skew"=round(isk,3), "kurtosi"=round(ik,3)))
           #round(outliersu,3), round(outliersl,3)))
}

# skip first column which is list of subsets
#datasett_index <- 1
#for (datasett in names(data_list)) {
  for (i in colnames(data_TPACK_pilot_minuspost)[colnames(data_TPACK_pilot_minuspost) %in% unlist(TPACK_list)]) {
    print(paste(i))
    describe_data(data_TPACK_pilot_minuspost[,i])
  }
#  datasett_index <- datasett_index + 1
#}

print(describe(data_TPACK_pilot_minuspost)) # independent variables, x
#print(describe(scoreItems(TPACK_list,data_TPACK_pilot_minuspost)$scores)) # dependent variables
```

@@

y = ax + b + restledd

Residuals should have an average value of 0, equal variance (homoscedasticity) and no interdependence (no autocorrelation) (Christophersen, 2018, kap. 7).

```{r residuals_singleitems}
check_residuals <- function(dataset, items) {

  table_residuals <- NULL

  teller <- 0                          # hvor mange kombinasjoner sjekker jeg?
  i      <- 1                          # start posisjon
  end    <- length(items)
  for (var1 in items[1:(end-1)]) {
    ifelse (length(items)>1, 
      y  <- rowMeans(dataset[, items]), # y-variabel
      y  <- dataset[, items])           # y består av bare ett item

    i    <- i + 1                      # for å unngå duplikater
    for (var2 in items[i:end]) {       # x-variabel
      x  <- dataset[,var2]
      xy <- data.frame(t(rbind(y,x)))
      colnames(xy) <- c(var1,var2)

      model <- lm( as.formula(
      paste( "xy [,'", var1, "'] ~ xy [,'", var2, "']", sep="")))

      print(c(var1, "=", var2))

      # check for outliers in residuals
      descriptive <- (describe_data(model$residuals)) # returns ("mean"=im, "skew"=isk, "kurtosi"=ik, outliersu, outliersl)

      # grense for betydningsfulle outliers / leverage; s. 78 i Christophersen, 2018
      k          <- length(items) # antall uavhengige variables
      n          <- nrow(xy)
      ifelse (k<10, 
        leverage <- 3*(k+1)/n,
        leverage <- 2*(k+1)/n
      )
      #print(c("betydningsfull grense:", leverage))

      # check for normalfordeling
      # 1: normalfordeling av residuals?
      hist(residuals(model), col = "steelblue")
      # 2: jevn fordeling av varians, homoskedastisitet
      print(plot(fitted(model), residuals(model)))
      print(abline(h = 0, lty = 2))

      teller <- teller + 1   # count number of scatterplots

      # check for autocorrelation
      dw     <- durbinWatsonTest(model)
      print(dw)

      table_residuals <- rbind(table_residuals, c("var1"=var1, "var2"=var2, 
                                                  descriptive, 
                                                  "dw=2"=round(dw$dw,2)))
    } 
  }
  print(teller) # antall plot

  # gjør om til tabell
  table_residuals         <- data.frame(table_residuals)

  # gjør om til tall
  table_residuals$mean    <- as.numeric(table_residuals$mean)
  table_residuals$skew    <- as.numeric(table_residuals$skew)
  table_residuals$kurtosi <- as.numeric(table_residuals$kurtosi)
  table_residuals$'dw.2'  <- as.numeric(table_residuals$'dw.2')

  return (table_residuals)
}

items     <- colnames(data_TPACK_pilot_minuspost)[colnames(data_TPACK_pilot_minuspost) %in% unlist(TPACK_list)]
table_res <- check_residuals(data_TPACK_pilot_minuspost, items)

# print items with large skew og kurtosi or autocollinearity
pander(table_res[abs(table_res$skew) > 1. | abs(table_res$kurtosi) > 3. | table_res$dw.2 > 2.5 | table_res$dw.2 < 1.5,])
```

Are there correlating residuals?  
Autocorrelation: Durbin-Watsons d = 2*(1-r)

If d = 2, then OK  
If d < 1 eller d > 3, then not OK

If autocorrelation, use robust SE

Autocorrelation can be due to structured data. Then use multilevel-analysis.

```{r distribution_residuals_aggregated}
teller         <- 0                             # hvor mange plot lager jeg?
datasett_index <- 1
#for (dataset in names(data_list)) {
#  datasett <- data_list[[dataset]]      # access data object
  for (var1 in names(TPACK_list)) {     # TPACK-element
    items      <- colnames(data_TPACK_pilot_minuspost)[colnames(data_TPACK_pilot_minuspost) %in% TPACK_list[[var1]]]
    ifelse (length(items)>1,            # ikke gjør analyse hvis bare ett item
      y        <- rowMeans(data_TPACK_pilot_minuspost[, items]), # y-variabel
      y        <- 0)

   print(c(var1, "=", items))

   if (!length(y)==1) {                # sjekk for at vi har flere enn et item

      # lag tekststreng til formel y= x1 + x2 + osv.
      independent_var <- paste("'] ~ xy [,'", items[1], sep="") #TK1"
      for (item in items[2:(length(items))]) { 
        independent_var <- paste(independent_var, "'] + xy [,'", item, sep="")
      }
      independent_var <- paste(independent_var, "'] - xy  [,'", sep="")

      #for (item in items) {
        model <- lm( as.formula(
        paste( "xy [,'", item, independent_var, item, "']", sep="")))

        # check for outliers in residuals
        print(describe_data(model$residuals))

        # grense for betydningsfulle outliers / leverage; s. 78 i Christophersen, 2018
        k <- length(items) # antall uavhengige variables
        n <- nrow(xy)
        ifelse (k<10, 
                leverage <- 3*(k+1)/n,
                leverage <- 2*(k+1)/n
                )
         print(c("betydningsfull grense:", leverage))

        # check for normalfordeling
        # 1: normalfordeling av residuals?
        print(hist(residuals(model), col = "steelblue"))
        # 2: jevn fordeling av varians, homoskedastisitet
        print(plot(fitted(model), residuals(model)))
        print(abline(h = 0, lty = 2))

        teller <- teller + 1   # count number of scatterplots
      #}
    }
  }
  #datasett_index <- datasett_index + 1
#} 
print(teller) # antall plot

@@

Missing data reduces the size of the dataset used for analysis, which leads to lower power (i.e., larger chances for Type II-error, that is, not detecting significant effects, obtaining false negatives). There are different advice for how to handle missing data, depending on whether they are missing at random (MAR), completely at random (MCAR) or not at random (NMAR) [@rubin_inference_1975; @schafer_missing_2002]. Data that are missing at random are correlated with the questionnaire (i.e., a question may be missing from a specific version of a questionnaire, which is in a sense random). Data that are missing completely at random, do however not correlate with anything obvious. They are in a sense, seemingly truly random, with no clear rationale for their missing other than circumstance. Data missing not at random are questions that correlate with the question itself (i.e., several respondents may avoid a specific question, i.e., about gender, there seems to be a rationale, that is not a random cause, for data missing for this specific item).

@@

```{r}
# ref: https://gabriellajg.github.io/EPSY-579-R-Cookbook-for-SEM/lavaan-lab-12-sem-for-missing-data.html
md.pattern(data_TPACK_pilot_minuspost)
```

n = 37  
9 are missing andre_fag  
16 are missing subject and gender
